Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Seminar April 27, CEDAR, Umeå}

### 2017-05-19 slides3.Rnw
touch

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

@@

<<lexis2>>=
load("data/vb0.rda")
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

@@

##load("data/vb.rda")  ####################################### NOTE!!!!!!!!!!!!!!!! ############################
##load("data/vb0.rda")
load("data/vb.rda")

tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")
x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "n",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE, ylim = c(28, 90))
axis(1, at = c(seq(1960, 1990, by = 10), 2002, 2014), las = 1)
axis(2, at = c(40, 65, 90), las = 2)
axis(4, at = c(40, 65, 90), las = 2)
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    ##abline(v = i, lwd = 2, col = "red")
    lines(c(i, i), c(40, 90), lwd = 2, col = "red")
}
for (i in seq(1960, 1985, by = 5) + (365 - 61) / 365){
    lines(c(i, i + 5), c(35, 40), col = "red", lty = 2, lwd = 2)
    lines(c(i, i), c(35, 40), col = "red", lty = 2, lwd = 2)
}
lines(c(1990, 1990) + (365 - 61)/365, c(28, 40), lwd = 2, lty = 2, col = "red")
lines(c(1990 + (365 - 61) / 365, 2002), c(28, 40), lwd = 2, lty = 2, col = "red")
for (i in 2002:2014){
    ##abline(v = i, lwd = 2, col = "blue")
    lines(c(i, i), c(40, 90), col = "blue", lwd = 2)
}
for (i in 2002:2013){
    lines(c(i, i), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
    lines(c(i, i+1), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
}
lines(c(2002, 2014), c(40, 40), col = "blue", lwd = 1.5)
lines(c(2002, 2014), c(90, 90), col = "blue", lwd = 1.5)
lines(c(1960, 2002), c(90, 90), lwd = 2, col = "red")
lines(c(1960, 2002), c(40, 40), lwd = 2, col = "red")
##abline(h = c(35, 40, 65, 90), lty = 2)
points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
     pch = ".", col = "red")
box()
@

@@

load("data/vb.rda")
tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))

@@

{\scriptsize
<<aggregate,eval = TRUE, echo = TRUE>>=
vb0Tab <- aggregate(vb0[, c("event", "exposure")], 
                   vb0[, c("sex", "age", "period", "cohort", 
                          "urban", "civst", "hisclass")], 
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)
nr <- NROW(vb0Tab)
@
}

@@

<<educ>>=
load("data/lisaold.rda")
educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
educ$period <- factor(educ$period)
educ <- educ[!is.na(educ$educ), ]
levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
educ$educ <- relevel(educ$educ, ref = "middle")
educ$civst <- relevel(educ$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban, 
             data = educ[educ$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban, 
             data = educ[educ$sex == "female", ])
cols <- c("darkgreen", "red", "black")
@

>>>>

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=5.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
## Load remote data:
con <- url("http://capa.ddb.umu.se/data/vb0.rda")
##load("data/vb0.rda")
load(con)
close(con)
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
##
con <- url("http://capa.ddb.umu.se/data/vb.rda")
##load("data/vb.rda")
load(con)
close(con)
##
con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
##load("data/lisaold.rda")
load(con)
close(con)
##
vb0Tab <- aggregate(vb0[, c("event", "exposure")],
                   vb0[, c("sex", "age", "period", "cohort",
                          "urban", "civst", "hisclass")],
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)

@@

<<lexis2>>=
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

@@

##load("data/vb.rda")  ####################################### NOTE!!!!!!!!!!!!!!!! ############################
##load("data/vb0.rda")
tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))
vbd <- vb0[vb0$event == 1 & vb0$dataset != "poplink", c("birthdate", "exit", "sex")]
d.m <- sum(vbd$sex == "male")
d.f <- sum(vbd$sex == "female")
x <- vbd$birthdate + vbd$exit
y <- vbd$exit
##par(mfrow = c(1, 2))
plot(x[vbd$sex == "male"], y[vbd$sex == "male"], type = "n",
     pch = ".", col = "blue", xlab = "Year", ylab = "Age", las = 2,
     axes = FALSE, ylim = c(28, 90))
axis(1, at = c(seq(1960, 1990, by = 10), 2002, 2014), las = 1)
axis(2, at = c(40, 65, 90), las = 2)
axis(4, at = c(40, 65, 90), las = 2)
for (i in seq(1960, 1990, by = 5) + (365 - 61) / 365){
    ##abline(v = i, lwd = 2, col = "red")
    lines(c(i, i), c(40, 90), lwd = 2, col = "red")
}
for (i in seq(1960, 1985, by = 5) + (365 - 61) / 365){
    lines(c(i, i + 5), c(35, 40), col = "red", lty = 2, lwd = 2)
    lines(c(i, i), c(35, 40), col = "red", lty = 2, lwd = 2)
}
lines(c(1990, 1990) + (365 - 61)/365, c(28, 40), lwd = 2, lty = 2, col = "red")
lines(c(1990 + (365 - 61) / 365, 2002), c(28, 40), lwd = 2, lty = 2, col = "red")
for (i in 2002:2014){
    ##abline(v = i, lwd = 2, col = "blue")
    lines(c(i, i), c(40, 90), col = "blue", lwd = 2)
}
for (i in 2002:2013){
    lines(c(i, i), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
    lines(c(i, i+1), c(39, 40), col = "blue", lty = 2, lwd = 1.5)
}
lines(c(2002, 2014), c(40, 40), col = "blue", lwd = 1.5)
lines(c(2002, 2014), c(90, 90), col = "blue", lwd = 1.5)
lines(c(1960, 2002), c(90, 90), lwd = 2, col = "red")
lines(c(1960, 2002), c(40, 40), lwd = 2, col = "red")
##abline(h = c(35, 40, 65, 90), lty = 2)
points(x[vbd$sex == "female"], y[vbd$sex == "female"], type = "n",
     pch = ".", col = "red")
box()
@

@@

##load("data/vb.rda")
tot.m <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "male"]))
tot.f <- length(unique(vb$id[vb$dataset != "poplink" & vb$sex == "female"]))

@@

{\scriptsize
<<aggregate,eval = TRUE, echo = TRUE>>=
nr <- NROW(vb0Tab)
@
}

@@

<<educ>>=

educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
educ$period <- factor(educ$period)
educ <- educ[!is.na(educ$educ), ]
levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
educ$educ <- relevel(educ$educ, ref = "middle")
educ$civst <- relevel(educ$civst, ref = "married")
fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "male", ])
fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
             data = educ[educ$sex == "female", ])
cols <- c("darkgreen", "red", "black")
@
